smdi: An R package to perform structural missing data investigations on partially observed confounders in real-world databases

### 2023-07-20 manuscript/manuscript.qmd
added section describing functions

```{r setup}
#| label: setup
#| include: false
library(magick)
library(cowplot)
library(here)
library(tidyverse)
library(readxl)
library(gt)
library(captioner)

table_cap <- captioner(prefix = "Table")
```

@@

In order to make an informed decision about the most appropriate analytic approach to arrive at unbiased estimates, it is crucial to investigate and understand the potential patterns and mechanisms that underlie the partially observed confounder data [@vanbuuren2018; @rubin1976; @little2019]. Usually these are not known for a given RWE study but general guidance papers and frameworks have suggested various routine diagnostics to investigate missing data patterns and mechanisms. These methods comprise standard procedures such as comparing distributions of baseline characteristics and outcomes between patients with and without the partially observed covariate(s) [@lee2021; @sondhi2023; @hotelling1931; @little1988], checking the ability to predict missingness based on observed data [@sondhi2023] and assessing if causal relationships between variables and their missingness are recoverable based on available data [@madley-dowd2019] and guided by directed acyclic graphs [@Lee2023; @Moreno-Betancur2018] or M-graphs [@mohan2021]. However, these methods have so far been only described and tested in isolation from each other and no principled approach exists. In addition, the practical implementation of all of these diagnostics is time-consuming, tedious and is consequently not often performed in both RWE and RCTs [@carroll2020; @wood2004; @harel2012].

@@

To streamline and ease the implementation of these routine missing data diagnostics for confounder data in RWE studies, we here present and demonstrate the use of the `smdi` (structural missing data investigations) R package [@smdi]. To that end, we first provide some brief background on the theoretical assumptions that underlie the diagnostic functions of this package. We then give an overview of the implemented functions and demonstrate an end-to-end workflow application of the package with a hypothetical RWE study using simulated oncology EHR dataset that comes bundled with the package.

@@

As opposed to clinical trials, which are designed to collect data for research purposes in a harmonized manner, real-world data (RWD) are typically generated for administrative purposes (e.g., health insurance claims for billing purposes) or clinical documentation (e.g., EHR). Hence, confounders and prognostic factors, which need to be balanced between treatment groups in RWE studies are usually not available for all patients at all neccessary time points [@toh2012a]. If the underlying mechanism for the missingness of such covariates is informative towards the outcome under study, e.g., patients who have a certain biomarker measured have a significantly increased risk if experiencing the outcome, this can lead to bias in the resulting effect estimates for the studied treatments under common missing data approaches like complete case analysis or imputation [@groenwold2020]. Hence, it is of utmost importance to investigate the potential patterns and mechanisms to know if assumptions for specific missing data approaches hold [@pedersen2017; @lee2013].

## Theoretical background

In our previous work [@weberpals2023] and for the purpose of the `smdi` R package, we categorized the implemented principled missing data investigations into three groups of diagnostics, which will be discussed in this section (Table 1).

Functions overview

Data generation for hypothetical study

@@

# Results

>>>>

```{r setup}
#| label: setup
#| include: false
library(magick)
library(cowplot)
library(here)
library(tidyverse)
library(readxl)
library(gt)
```

@@

In order to make an informed decision about the most appropriate analytic approach to arrive at unbiased estimates, it is crucial to investigate and understand the potential patterns and mechanisms that underlie the partially observed confounder data [@vanbuuren2018; @rubin1976; @little2019]. Usually these are not known for a given RWE study but general guidance papers and frameworks have suggested various routine diagnostics to investigate missing data patterns and mechanisms. These methods comprise standard procedures such as comparing distributions of baseline characteristics and outcomes between patients with and without the partially observed covariate(s) [@lee2021; @sondhi2023; @hotelling1931; @little1988; @pedersen2017], checking the ability to predict missingness based on observed data [@sondhi2023] and assessing if causal relationships between variables and their missingness are recoverable based on available data [@madley-dowd2019] guided by directed acyclic graphs [@Lee2023; @Moreno-Betancur2018] or M-graphs [@mohan2021]. However, these methods have so far been only described and tested in isolation from each other and no principled approach exists. In addition, the practical implementation of all of these diagnostics is time-consuming, tedious and is consequently not often performed in both RWE and RCTs [@carroll2020; @wood2004; @harel2012].

@@

To streamline and ease the implementation of these routine missing data diagnostics for confounder data in RWE studies, we here present and demonstrate the use of the `smdi` (structural missing data investigations) R package [@smdi]. To that end, we first provide some brief background on the theoretical assumptions that underlie the diagnostic functions of this package. We then give an overview of the implemented functions and demonstrate an end-to-end workflow application of the package with a hypothetical RWE study using a simulated oncology EHR dataset that comes bundled with the package.

@@

As opposed to clinical trials, which are designed to collect data for research purposes in a harmonized manner, real-world data (RWD) are typically generated for administrative purposes (e.g., health insurance claims for billing purposes) or clinical documentation (e.g., EHR). Hence, confounders and prognostic factors, which need to be balanced between treatment groups in RWE studies, are usually not available for all patients at all necessary time points [@toh2012a]. If the underlying mechanism for the missingness of such covariates is not at random, e.g., patients with higher levels of a prognostic biomarker are more likely to be missing, this can lead to bias in the resulting effect estimates for the studied treatments under common missing data approaches like complete case analysis or imputation [@groenwold2020]. Hence, it is of utmost importance to investigate the potential patterns and mechanisms to know if assumptions for specific missing data approaches hold [@pedersen2017; @lee2013].

## Theoretical background and main package functions

For the implementation of these routine diagnostics checks, we categorized the main functions of the `smdi` R package into three group diagnostics based on their general analytic purpose (Table 1). In this section, we will explain the basic ideas behind the main package functions, the parameters a user can specify, the results that are being returned by calling the functions and how the results can be interpreted. Generally, for all functions in the `smdi` package, a *dataframe* is expected as input with a format where one row stands for one unqiue patient and the columns stand for the variables relevant for the study, i.e., the exposure, outcome, fully observed covariates and the partially observed covariates (one-row-per-patient format). Any non-informative columns, such as patient identifiers, date columns or zip codes should be dropped from the dataframe before calling the function. Throughout all functions, the user has the option to specify a vector with the column names of the partially observed covariate(s) that should be investigated (`covar` parameter). If nothing is specified, all functions automatically consider any variable in the dataframe that exhibits at least one missing value.

### Group 1 diagnostics

The aim of the `smdi_asmd()`, `smdi_hotelling()` and `smdi_little()` functions is to explore dissimilarities in patient characteristics between those with and without observed values for the partially observed covariate. According to Rubin's framework on inference and missing data [@rubin1976], when missingness is at random (MAR), it can be explained by observed covariates. Consequently, significant differences in patient characteristics would be expected between strata of patients with and without the partially observed covariate. If the missingness depends only on unobserved factors (missing not at random \[MNAR\]) or does not depend on either observed or unobserved covariates (missing completely at random \[MCAR\]), differences should not be observable.

To quantify and test such differences, the `smdi_asmd()` function computes absolute standardized mean differences (ASMD) of observed patient characteristics in the dataset [@schober2019; @austin2011; @tableone]. To call the function, a user must specify the dataset th

function returns an *amsd* object which displays an aggregated summary of the average or median (depending on the analyst's choice) and the minimum/maximum range of the derived ASMDs. The *amsd* object also returns a detailed 'Table 1' for each partially observed covariate displaying the distributions of observed patient characteristics and resulting ASMDs between patients with and without an observed value for the partially observed covariate. For a graphical visualization, the function also creates a *ggplot2* graph [@ggplot2] illustrating the ASMDs for each compared patient characteristic in descending order.

The `smdi_hotelling()` and `smdi_little()` functions complement the `smdi_asmd()` function by examining the differences in patient characteristics as a formal statistical hypothesis test. Hotelling's test [@hotelling1931; @Hotelling] formalizes this is a multivariate t-test for each partially observed covariate, which means that `smdi_hotelling()` returns a test statistic and a corresponding p-value for each each partially observed covariate individually. In contrast, Little's test [@little1988; @naniar] (`smdi_little()`) computes a single global chi-square test statistic and p-value across all partially observed covariates with the null hypothesis that the data is MCAR.

### Group 2 diagnostics

The group 2 diagnostics assesses the ability to predict missingness based on observed covariates via the `smdi_rf()` functions. This function trains and fits and random forest classification model [@randomForest; @sondhi2023a] to predict the missing indicator of the partially observed covariate given observed covariates as the predictors. If the resulting area under the receiver operating characteristic curve (AUC) is meaningfully higher than 0.5, this would give some evidence for MAR being the underlying missingness mechanism. In case of values close 0.5, this would rather indicate a random prediction and translate to a potential MCAR or MNAR mechanism. The function returns an object of class `rf` which generically displays an overview of the AUC values of all partially observed covariates. For more details, the `rf` object additionally includes a *ggplot2* graph for each partially observed covariate displaying the relative importance of the predictors in the training dataset expressed as the mean decrease in accuracy. This metric can be valuable for interpreting and identifying strong predictors of missingness. It quantifies how much the accuracy of the prediction (i.e., the ratio of correct predictions to the total number of predictions made) would decrease if we excluded a specific predictor from the prediction model. For tuning the random forest model, the `smdi_rf()` function has a few parameters that can be specified by the user for optimization such as the number of trees to grow (default is 1,000 trees), the ratio of the size between train and test datasets (default is a 70/30 split) and the number of cores to parallelize the computation on [@parallel], since this function may be very time consuming, especially with larger datasets.

### Group 3 diagnostics

Finally, the third group diagnostics with the `smdi_outcome()` function examines association of the missingness indicator of the partially observed covariate and the outcome under study. The function will compute both a univariate model and a model adjusted for all other covariates specified in the dataset. Currently, three outcome regression types are supported: linear regression (*lm* [@stats]) for continuous outcomes, logistic regression (*glm* [@stats]) for binary outcomes and a Cox proportional hazards model (*coxph* [@survival]) for time-to-event outcomes. Besides the specification of the regression type, the user needs to specify the name of the column that contains the outcome using the *form_lhs* parameter (e.g., `Surv(eventtime, status)` in case of a Cox model) and if resulting beta coefficients should be displayed exponentiated or not.

## Data generation for hypothetical study

@@

# Results  {#sec-results}
